\section{Synthesis: The Conformance Functor}
\label{sec:synthesis-functor}

The five pillars converge on a single object. Pillar~I gave categories and the
monoidal product; Pillar~II gave the verdict lattice and monotonicity; Pillars~III
and~IV gave the transition and the manifold; Pillar~V gave the measure. We now show
that conformance-for-all-language is one \emph{functor}, and that the thesis's three
load-bearing results---non-collapse, refusal-survives-composition, and
admission-carries-measure---are its three defining properties.

\begin{definition}[Event category and verdict category]
\label{def:event-cat}
For a domain $d$ (\cref{ch:all-language}), let the \emph{event category}
$\mathcal E_d$ have as objects the object-centric event logs over $d$'s schema and as
morphisms the \emph{evidence refinements} (log extensions that add events or resolve
attributes without contradicting existing order). Let the \emph{verdict category}
$\mathcal L$ be the verdict lattice of \cref{def:verdict} regarded as a thin
category: objects are verdict assignments and there is a unique morphism
$v\to w$ exactly when $v\sqsubseteq w$ in the knowledge order.
\end{definition}

\begin{theorem}[The conformance functor]
\label{thm:conf-functor}
Fix for each domain $d$ a model $M_d$, and let $\mathrm{Conf}_d\colon\mathcal E_d\to
\mathcal L$ send a log to its verdict against $M_d$ (\cref{def:conformance-surface}).
Then:
\begin{enumerate}[label=(\roman*),leftmargin=2.2em]
  \item each $\mathrm{Conf}_d$ is a functor;
  \item the domains assemble into the coproduct category
  $\mathcal E=\coprod_d\mathcal E_d$, and the copairing
  $\mathrm{Conf}=[\mathrm{Conf}_d]_d\colon\mathcal E\to\mathcal L$ is a single
  functor---one conformance surface for all language;
  \item $\mathrm{Conf}$ is monotone (\cref{thm:noncollapse}): it sends evidence
  refinements to knowledge increases and never to collapses.
\end{enumerate}
\end{theorem}
\begin{proof}
(i) A functor must preserve identities and composition. The identity refinement
(adding nothing) leaves the verdict fixed, and $\mathrm{Conf}_d$ of a composite
refinement equals the composite of verdict morphisms because verdict depends only on
the accumulated evidence, which composes associatively (\cref{prop:replay-lfp}).
Functoriality requires that a refinement $\ell\to\ell'$ map to a morphism
$\mathrm{Conf}_d(\ell)\to\mathrm{Conf}_d(\ell')$ in $\mathcal L$, i.e.\ to a
knowledge increase $\sqsubseteq$; this is exactly the monotonicity of
\cref{thm:noncollapse}. (ii) The coproduct $\mathcal E$ has the universal property
that a functor out of it is precisely a family of functors out of the summands; the
copairing $[\mathrm{Conf}_d]_d$ is that functor, defined on the summand $\mathcal
E_d$ as $\mathrm{Conf}_d$. (iii) is (i) read across all summands.
\end{proof}

\begin{figure}[t]
\centering
\begin{tikzcd}[ampersand replacement=\&,column sep=large]
\mathcal E_d \arrow[r,hook,"\iota_d"] \arrow[dr,"\mathrm{Conf}_d"'] \& \mathcal E \arrow[d,"\mathrm{Conf}"]\\
\& \mathcal L
\end{tikzcd}
\caption[The conformance functor]{Each domain's conformance $\mathrm{Conf}_d$
factors through the one functor $\mathrm{Conf}$ on the coproduct of event
categories (\cref{thm:conf-functor}). ``All language, one surface'' is the
commutativity of this triangle for every domain $d$.}
\label{fig:conf-functor}
\end{figure}

The thesis's two further results are now properties of this functor.

\begin{proposition}[Refusal survives composition]
\label{prop:refusal-survives}
Equip $\mathcal E$ with the monoidal product $\otimes$ that fuses two logs
(Pillar~I) and $\mathcal L$ with the axiswise truth-order meet $\wedge_t$. Then
$\mathrm{Conf}(x\otimes y)\preceq_t\mathrm{Conf}(x)\wedge_t\mathrm{Conf}(y)$: on each
axis the fused verdict is no more admitted than the meet of the parts. In
particular a \refused\ axis in either part is \refused\ in the fusion.
\end{proposition}
\begin{proof}
On an axis $a$, the fused log contains the evidence of both parts. If either part
refuses $a$ (witnessing a violation), the violation persists in the fusion, so the
fused verdict is $\mathsf f=\min_t$. Admission ($\mathsf t$) of the fusion requires
the violation-free evidence of both, hence both parts admit; and an axis
\unknownstat\ in either part cannot exceed $\mathsf u$ in the fusion without new
evidence (\cref{thm:noncollapse}). In every case the fused value is
$\le_t\min_t(\mathrm{Conf}(x)_a,\mathrm{Conf}(y)_a)$.
\end{proof}

\Cref{prop:refusal-survives} is the compositor invariant of \cref{sec:fus-stack}
(``\code{REFUSED\_BY\_LAW} codes survive always'') stated as a lax-monoidality, and
it is the formal guarantee behind the portable verdict of
\cref{sec:fus-conformance}: fusing more agents can only lower, never launder, an
admission.

\begin{proposition}[Admission carries measure]
\label{prop:admission-measure}
An \admitted\ verdict of $\mathrm{Conf}$ on a trajectory is accompanied by positive
receipt measure: $\mu_R(\{\gamma\})>0$ (\cref{prop:receipt-measure}). Equivalently,
$\mathrm{Conf}$ factors through the witnessed set $W$, and admission off $W$ is not
in the image.
\end{proposition}
\begin{proof}
Admission requires a receipt by \cref{def:lawstate}; a receipt assigns positive
weight $w(\gamma)>0$, so $\mu_R(\{\gamma\})>0$ by \cref{con:receipt-measure}.
Trajectories with $w=0$ lie outside $W$ and cannot be admitted, so $\mathrm{Conf}$
restricted to admission has image in the witnessed set.
\end{proof}

\begin{corollary}[All language, one functor]
\label{cor:one-functor}
The seven domains of \cref{ch:all-language}---natural language, legal text,
mathematics, biological sequence, music, business process, and the reflexive case
of an agent's own conduct---are objects of $\mathcal E$, and a single functor
$\mathrm{Conf}$ (\cref{thm:conf-functor}), monotone (\cref{thm:noncollapse}),
lax-monoidal (\cref{prop:refusal-survives}), and measure-valued
(\cref{prop:admission-measure}), is their common conformance surface. Passing from
one domain to another changes the object and the model $M_d$; it does not change the
functor.
\end{corollary}

\Cref{cor:one-functor} is the mathematical statement of the thesis. The reach of the
fused protocols (Pillar~I) and the trustworthiness of three-valued, receipt-bound,
measure-carrying conformance (Pillars~II and~V) are not two systems bolted together
but one functor on one category---and the phase transition of
Pillars~III and~IV is what carries that functor from the single domain where van der
Aalst built it to the whole of language.
