\section{Pillar IV --- The Geometry of Law-State}
\label{sec:geometry}

The artifact's constitution closes with a striking passage: ``Repo state is the
manifold. Agent edits are trajectories. Failsets are curvature. Receipts are proof
measure. LSP is the differential operator. Diagnostics are gradients. Code actions
are constrained control vectors. Admissibility is $\Phi_G=0$''
\citep{lspmax}. This pillar takes that passage literally and builds, from
topological first principles, the differential geometry that makes each line a
definition rather than a flourish.

\subsection{From topology to smooth manifolds}
\label{sec:geo-manifold}

\begin{definition}[Topological space, continuity]
A \emph{topology} on a set $X$ is a collection $\tau$ of subsets (the
\emph{opens}) containing $\varnothing$ and $X$ and closed under arbitrary unions
and finite intersections. A map $f\colon X\to Y$ is \emph{continuous} if
preimages of opens are open. $X$ is \emph{Hausdorff} if distinct points have
disjoint open neighbourhoods.
\end{definition}

\begin{definition}[Smooth manifold]
An \emph{$n$-dimensional smooth manifold} $M$ is a second-countable Hausdorff
space with an \emph{atlas} of \emph{charts} $\varphi_\alpha\colon U_\alpha\to
\mathbb R^{n}$ (homeomorphisms from opens $U_\alpha$ covering $M$) whose
transitions $\varphi_\beta\circ\varphi_\alpha^{-1}$ are smooth ($C^\infty$) where
defined. A map between manifolds is \emph{smooth} if it is smooth in charts.
\end{definition}

\begin{notation}[The state manifold]
We model the repository's admissible configuration space as a smooth manifold $M$:
points are configurations, and a chart is a local coordinate system on
configurations (in practice, a continuous relaxation of edit space). This is the
sense of ``repo state is the manifold''; the idealisation to a smooth $M$ is what
licenses the calculus below, and is itself a modelling choice we flag as
\candidate.
\end{notation}

\subsection{Tangent vectors, fields, and trajectories}
\label{sec:geo-tangent}

\begin{definition}[Tangent space]
A \emph{tangent vector} at $p\in M$ is a derivation $v\colon C^\infty(M)\to\mathbb
R$: a linear map with $v(fg)=v(f)\,g(p)+f(p)\,v(g)$ (the Leibniz rule). The
tangent vectors at $p$ form the \emph{tangent space} $T_pM$, an $n$-dimensional
vector space; their disjoint union is the \emph{tangent bundle} $TM$. A
\emph{vector field} is a smooth section $X\colon M\to TM$.
\end{definition}

\begin{definition}[Trajectory]
A \emph{trajectory} (integral curve) of a vector field $X$ is a smooth
$\gamma\colon I\to M$ with $\dot\gamma(t)=X(\gamma(t))$. An agent's sequence of
edits is modelled as such a curve, and a \emph{code action} is the control vector
$\dot\gamma(t)\in T_{\gamma(t)}M$ it applies---``agent edits are trajectories''
and ``code actions are constrained control vectors''.
\end{definition}

\subsection{Metric, gradient, and the descent of failure}
\label{sec:geo-metric}

\begin{definition}[Riemannian metric and gradient]
A \emph{Riemannian metric} $g$ assigns to each $p$ an inner product
$g_p(\cdot,\cdot)$ on $T_pM$, smoothly in $p$. For $f\in C^\infty(M)$ the
\emph{gradient} $\nabla f$ is the unique vector field with
$g_p(\nabla f, v)=\mathrm df_p(v)$ for all $v\in T_pM$, where $\mathrm df$ is the
differential. \emph{Gradient flow} is the trajectory of $-\nabla f$.
\end{definition}

Let $\Phi_G\colon M\to\mathbb R_{\ge 0}$ be the \emph{failset potential}: a smooth
penalty aggregating the active law-axis violations (the continuous shadow of the
\code{ConformanceVector}, zero exactly on configurations whose active axes are all admitted). Then
``diagnostics are gradients'' is the identification of the emitted corrective
direction with $-\nabla\Phi_G$, and ``LSP is the differential operator'' is the
reading of the read-only LSP surface as the map $f\mapsto\mathrm df$ that exposes
that gradient without itself moving the point (\cref{sec:art-overview}).

\begin{proposition}[Admissibility descends]
\label{prop:descent}
Along the gradient flow $\dot x=-\nabla\Phi_G(x)$, the potential is
non-increasing: $\tfrac{\mathrm d}{\mathrm dt}\Phi_G(x(t))=-\lVert\nabla\Phi_G
\rVert_g^{2}\le 0$, with equality only at critical points. Hence trajectories move
monotonically toward the admissible region.
\end{proposition}
\begin{proof}
By the chain rule and the defining property of the gradient,
$\frac{\mathrm d}{\mathrm dt}\Phi_G(x)=\mathrm d\Phi_G(\dot x)
=g(\nabla\Phi_G,-\nabla\Phi_G)=-\lVert\nabla\Phi_G\rVert_g^{2}\le 0$,
which vanishes iff $\nabla\Phi_G=0$.
\end{proof}

\Cref{prop:descent} gives ``effective admitted velocity $\mathrm
dA_{\text{admitted}}/\mathrm dt$'' a precise reading: the rate at which a
trajectory reduces $\Phi_G$, maximised by following the diagnostic gradient. The
artifact's slogan ``optimise for admitted work, not raw output'' is the
instruction to descend $\Phi_G$ rather than to maximise path length in $M$.

\subsection{The admissible submanifold (admissibility as \texorpdfstring{$\Phi_G=0$}{Phi=0})}
\label{sec:geo-submanifold}

\begin{theorem}[Regular value / preimage theorem]
\label{thm:regval}
Let $F\colon M\to\mathbb R^{k}$ be smooth and let $0$ be a \emph{regular value}:
$\mathrm dF_p$ is surjective at every $p\in F^{-1}(0)$. Then $A=F^{-1}(0)$ is a
smooth submanifold of $M$ of codimension $k$, with $T_pA=\ker\mathrm dF_p$.
\end{theorem}
\begin{proof}[Proof sketch]
At $p\in A$, surjectivity of $\mathrm dF_p$ lets one choose coordinates in which
$F$ is a projection onto the last $k$ coordinates (implicit function theorem); in
those coordinates $A$ is the coordinate slice where they vanish, a submanifold of
codimension $k$, and its tangent space is the kernel of the projection's
differential, $\ker\mathrm dF_p$.
\end{proof}

\begin{corollary}[Geometry of admissibility]
\label{cor:admissible}
Write $\Phi_G=(\phi_1,\dots,\phi_k)$ for the $k$ independent active law-axis
penalties. If $0$ is a regular value, the admissible set
$A=\{\,x:\Phi_G(x)=0\,\}$ is a submanifold of codimension $k$: each independent
law removes one degree of freedom. Admissible trajectories are exactly those
remaining within $A$, i.e.\ with control vectors in $T_xA=\ker\mathrm d\Phi_{G,x}$.
\end{corollary}

\Cref{cor:admissible} makes ``admissibility is $\Phi_G=0$'' exact and quantifies
it: the admissible world is not a point but a submanifold whose codimension counts
the binding laws, and lawful action is motion tangent to it. The three-valued
discipline of Pillar~II reappears geometrically: a point off $A$ is \refused\
(some $\phi_i>0$), a point on $A$ with verified evidence is \admitted, and a
direction not yet known to lie in $T_xA$ is \unknownstat\ until checked.

\subsection{Connection, curvature, and ``failsets are curvature''}
\label{sec:geo-curvature}

\begin{definition}[Affine connection, geodesic, curvature]
An \emph{affine connection} $\nabla$ assigns to vector fields $X,Y$ a field
$\nabla_XY$, $\mathbb R$-linear in $Y$, $C^\infty(M)$-linear in $X$, and Leibniz
in $Y$. A \emph{geodesic} satisfies $\nabla_{\dot\gamma}\dot\gamma=0$
(``straightest'' curves). The \emph{Riemann curvature} is
$R(X,Y)Z=\nabla_X\nabla_YZ-\nabla_Y\nabla_XZ-\nabla_{[X,Y]}Z$, measuring the
failure of second covariant derivatives to commute.
\end{definition}

Curvature measures how constraints bend straightness. We propose the precise
reading of ``failsets are curvature'': near the admissible submanifold $A$, the
\emph{second fundamental form} (the normal component of $\nabla_XY$ for $X,Y$
tangent to $A$) records how the failset potential $\Phi_G$ curves the admissible
sheet, equivalently the Hessian $\operatorname{Hess}\Phi_G$ restricted to the
normal bundle. Where this curvature is large, small lawful steps demand large
corrective detours; where it is flat, admissible motion is cheap. The failset is
not a wall but a shape, and its shape is curvature.

\begin{table}[t]
\centering
\caption[The geometry of law-state]{A precise reading of the artifact's manifold
metaphors \citep{lspmax}. Each informal line on the left is assigned a geometric
object derived in this pillar.}
\label{tab:geometry}
\small
\begin{tabularx}{\textwidth}{@{}l Y@{}}
\toprule
\textbf{\code{AGENTS.md} metaphor} & \textbf{Geometric object (this pillar)} \\
\midrule
Repository state is the manifold & smooth manifold $M$ (\cref{sec:geo-manifold}) \\
Agent edits are trajectories & integral curves $\gamma$ of a vector field (\cref{sec:geo-tangent}) \\
Code actions are constrained control vectors & tangent vectors $\dot\gamma\in T_xA$ (\cref{cor:admissible}) \\
LSP is the differential operator & the map $f\mapsto\mathrm df$ (read-only) (\cref{sec:geo-metric}) \\
Diagnostics are gradients & $-\nabla\Phi_G$ (\cref{prop:descent}) \\
Admissibility is $\Phi_G=0$ & the submanifold $A=\Phi_G^{-1}(0)$ (\cref{cor:admissible}) \\
Failsets are curvature & second fundamental form / $\operatorname{Hess}\Phi_G|_{NA}$ (\cref{sec:geo-curvature}) \\
Effective velocity $\mathrm dA_{\text{admitted}}/\mathrm dt$ & descent rate $-\lVert\nabla\Phi_G\rVert^2$ (\cref{prop:descent}) \\
Receipts are proof measure & a measure on trajectory space (\cref{sec:geo-measure}) \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Receipts as proof measure}
\label{sec:geo-measure}

The final metaphor, ``receipts are proof measure'', asks for measure theory. A
\emph{measure} assigns non-negative, countably additive mass to (measurable) sets;
a receipt ledger (\cref{sec:art-receipts}) assigns admitted mass to the set of
trajectories it witnesses. Admissible volume is then $\mu(A\cap\,\cdot\,)$ rather
than path length: the artifact's refusal to equate ``ran far'' with ``did
admitted work'' is the refusal to confuse Lebesgue length of $\gamma$ with the
receipt-measure of the admissible region it traversed. A trajectory may be long
and carry zero proof measure; a short one, bound to a valid receipt chain, carries
positive measure. This is \cref{thm:noncollapse} once more, now measure-theoretic:
unwitnessed motion contributes no admitted mass, just as unevidenced axes never
leave \unknownstat.

\medskip
Four pillars---composition, verdicts, transition, and law-state---are now in place.
A fifth (\cref{sec:measure}) supplies the measure and information theory that the
artifact's ``receipts are proof measure'' demands, after which a short synthesis
(\cref{sec:synthesis-functor}) draws the five into a single functor.
