\section{Recent Advances (December 2025 -- June 2026)}
\label{sec:recent}

The thesis is written into a fast-moving literature. This section surveys
seventy preprints from the six months preceding submission to show that the
argument sits at a live confluence rather than on settled ground.\footnote{The
works surveyed here were located by a machine-assisted arXiv harvest restricted to
the window 2025-12-01 to 2026-06-21; fifteen were additionally verified by hand
against their arXiv abstract pages (identifier, title, authors, and v1 date), and
the remainder by the harvest's per-item fetch. As living preprints whose contents
and versions change, every entry should be re-verified before formal submission;
the set carries bounded status \candidate\ rather than \admitted.} The survey is
organised by the five research currents the thesis joins.

\paragraph{Process mining is becoming agentic and object-centric.} The discipline
of \cref{sec:bg-pm} is, in real time, turning toward the agents and the
object-centric data this thesis foregrounds. Van der Aalst's own group has
released \emph{PMAx}, ``an agentic framework for AI-driven process mining'' that
turns natural-language business questions into mining artifacts
\citep{pmax26}, and a neuro-symbolic anomaly detector that injects \emph{Declare}
constraints into neural conformance \citep{gaikwad26}---the very Declare-plus-OCEL
combination the artifact of \cref{ch:artifact} uses. Object-centric event data is
receiving fresh ontological foundations for time and relations \citep{hooshyar25}
and comparative grounding against industrial multilevel mining \citep{ronzoni25}.
LLMs are entering predictive monitoring \citep{padella26}, business-process
modelling assessment \citep{lauer26}, multi-agent work-in-progress prediction
\citep{bibalan25}, and clinical-pathway risk estimation \citep{ardimento26}, while
the field reflects on ``process analytics'' as a discipline \citep{stierle25},
explainable workflow verification \citep{klimek25}, decentralised edge mining
\citep{reiter25}, and network-traffic mining \citep{vitale25}. The thesis's
premise---that conformance over object-centric logs is the bridge to AI---is the
premise the field itself is now adopting.

\paragraph{Agent interoperability is consolidating around MCP and A2A.} The
landscape of \cref{sec:bg-landscape} is maturing fastest along its security and
semantics. A systematisation of MCP security \citep{gaire25} sits beside
spec-level analyses \citep{maloyan26}, tool-poisoning attacks \citep{li26mcpitp,
huang26threat}, defense-placement taxonomies \citep{rostamzadeh26}, and verifiably
safe tool use \citep{doshi26}; orchestration is surveyed \citep{adimulam26} and
made task-adaptive \citep{yu26adaptorch}, trained on traces \citep{zhang26rl}, and
re-examined as design choice \citep{felendler26} and workflow engine
\citep{parmar26}. Most relevant to \cref{hyp:substrate}, a wave of work pushes
\emph{beyond} message passing toward the conformance-bearing strata this thesis
argues for: a ``semantic view'' of agent communication protocols
\citep{yuan26semantic}, identity and verifiable delegation \emph{across MCP and
A2A} \citep{prakash26aip, patil26sentinel, otsuka26}, cryptographic capability
binding with reproducibility \citep{zhou26gov}, new delegation and memory
protocols \citep{nie26, xu26mesh}, and agentic peer-to-peer substrates
\citep{wang26p2p}. The cryptographic-binding and verifiable-delegation works are,
in effect, independent rediscoveries of the receipt discipline of
\cref{sec:art-receipts}.

\paragraph{Agentic AI is preoccupied with reliability, evaluation, and
guardrails.} The deficiency this thesis names P2---reach without
accountability---is the field's dominant anxiety. There are calls for a ``science
of AI agent reliability'' \citep{rabanser26} and unified evaluation
\citep{zhu26eval, souza26, arunkumar26}, benchmarks for MCP tool use
\citep{mcpagentbench25}, productivity-agent safety \citep{li26claws}, and test-time
scaling \citep{li26scaling}; diagnoses of tool-invocation failure
\citep{huang26fail} and agentic-framework bugs \citep{zhang26bug}; and a
proliferation of guardrail and oversight mechanisms---learned access control
\citep{abaev26}, diagnostic guardrails \citep{liu26dog}, proof-of-guardrail
\citep{jin26guardrail}, provably secure guardrails \citep{wu26guardrail},
self-healing \citep{jeong26}, and the sobering observation that human oversight
has finite capacity \citep{turan26}. Agent memory and ``skills'' are being
formalised and governed \citep{du26memory, lam26memory, xu26skills, zhou26ext}.
Each of these is a partial, domain-specific instance of the law-state surface this
thesis argues should be a first-class, three-valued, receipt-bound stratum
(\cref{prin:three-valued}); ``proof-of-guardrail'' \citep{jin26guardrail} is the
phrase the field reaches for when it wants exactly a receipt.

\paragraph{Compositional and formal foundations remain active.} Pillar~I's
categorical framing (\cref{sec:alg}) is contemporary, not antiquarian. Recent work
develops string-diagrammatic and categorical accounts of stochastic and
non-deterministic systems \citep{bonchi26, lynch26}, profunctorial algebra
\citep{aristote26}, monadic metalanguages \citep{liellcock25}, and categorical
semantics for real systems \citep{carlisle26}; abstract interpretation
\citep{lueker25}, dataflow analysis \citep{brahmakshatriya26}, session-typed
metaprogramming \citep{angelo26}, and game-semantic verification
\citep{koutavas25} keep the program-analysis lineage of LSP active. Categorical
systems theory \citep{lynch26} is, in particular, the natural setting for the
operadic reading of agent composition in \cref{sec:alg-operad}.

\paragraph{Learning itself is being read as a phase transition.} Pillars~III and
IV (\cref{sec:analysis,sec:geometry}) keep close company with a surge of work
treating deep learning through statistical mechanics and geometry. Phase
transitions are reported in feature learning \citep{montanari26}, in the
hierarchical structure of trained networks \citep{ersoy25}, and as ``cascades''
in scaling \citep{defilippis26}; ``grokking'' is analysed as a finite-size
transition \citep{bi26grok}, a dimensional transition \citep{wang26grok}, a
metastable escape \citep{ersoy26}, and over non-commutative algebra
\citep{notsawo26}. The geometry of representations and generalisation
\citep{yadav26}, Riemannian approximation \citep{david25}, and
renormalisation-group methods \citep{tan26rg} are all under active study, and a
fourteen-author manifesto argues outright that ``there will be a scientific theory
of deep learning'' built on solvable limits, scaling laws, and universality
\citep{simon26}. That learning is now routinely modelled with the precise tools
this thesis applies to protocol convergence---phase transitions, latent variables,
curvature---is the strongest external evidence that the framing of
\cref{ch:framework} is the right register, and the company the model of
\cref{sec:an-standardisation} keeps.

\medskip
\noindent Two cautions close the survey. First, half-life: a six-month citation
window is, by construction, perishable, and several of these preprints will be
superseded or revised. Second, direction of fit: the convergence of these
literatures toward conformance, receipts, composition, and transition is
consistent with this thesis's claims but does not by itself confirm them; it
establishes timeliness, not truth. The confirmation, such as it is, is the
argument of the chapters that follow.
