\chapter{The \texorpdfstring{\lspmax}{lsp-max} Artifact}
\label{ch:artifact}

\epigraph{The repository is not passive text. It is a law-bearing
system.}{\textit{\lspmax}, \texttt{AGENTS.md}}

This chapter presents \lspmax\ as the instantiation of the law-state runtime of
\cref{def:lawstate}. It is, in the words of its own manifest, ``a law-state
runtime projected through LSP'' \citep{lspmax}, and it is studied here at
CalVer~26.6.18. The chapter traces its five-layer architecture
(\cref{sec:art-layers}), its \maxm{*} protocol surface (\cref{sec:art-maxm}), its
three-valued \code{ConformanceVector} (\cref{sec:art-cv}), its receipts and
typestate kernel (\cref{sec:art-receipts}), its self-enforcing constitution
(\cref{sec:art-canary}), and---the point of contact with van der Aalst's
formalisms---its binding of diagnostics to object-centric events
(\cref{sec:art-ocel}).

\section{Overview: Inverted LSP}
\label{sec:art-overview}

A conventional language server helps a human write code. \lspmax\ inverts the
direction: it ``makes the repository speak back to agents while they work''
\citep{lspmax}. The same JSON-RPC surface that an editor uses to ask ``what does
this symbol mean?'' is repurposed so that an agent may ask ``is this trajectory
admissible?''. The artifact's constitution, \code{AGENTS.md}, states the governing
condition for admission in a form worth reproducing, because it is
\cref{def:lawstate} in the artifact's own notation:
\[
  \mathrm{Do{n}e}_B(A) \;=\; \big[\,\mathrm{FailSet}_B(A)=\varnothing\,\big]
  \;\wedge\; \big[\,R_B \vdash A = \mu(O^{*}_B)\,\big].
\]
In words: agent output is admitted not because it compiles, logs success, or
passes a test, but only when bounded receipts prove the action equals the lawful
transformation of admitted observations \citep{lspmax}. The artifact thus refuses
exactly the inference---``the test passed, therefore the work is admissible''---that
\cref{prin:three-valued} forbids.

\section{The Five-Layer Model}
\label{sec:art-layers}

\lspmax\ is organised as five layers in which ``law-state'' is the load-bearing
abstraction (\code{docs/ARCHITECTURE.md}). \Cref{tab:five-layers} lists them with
the crates that realise each.

\begin{table}[t]
\centering
\caption[The five-layer model of \lspmax]{The five-layer model and its crate
mapping \citep{lspmax}. Layer numbering runs from the actuation grammar (1) up to
the autonomic mesh (5); ``law-state'' (Layer 3) is the load-bearing abstraction.}
\label{tab:five-layers}
\small
\begin{tabularx}{\textwidth}{@{}c l Y@{}}
\toprule
\textbf{Layer} & \textbf{Name} & \textbf{Realising crate(s)} \\
\midrule
5 & Autonomic LSP mesh & \code{lsp-max-compositor}; example servers \\
4 & Knowledge hooks & \code{src/composition/}; hook graph \\
3 & Law-state runtime & \code{lsp-max-runtime}, \code{lsp-max-protocol} \\
2 & Local LSP state surface & root crate \code{src/} (\code{LanguageServer}, \code{LspService}, \code{Server}) \\
1 & Actuation grammar & \code{crates/lsp-max-cli} (clap-noun-verb) \\
\bottomrule
\end{tabularx}
\end{table}

Two design rules of the model matter for the thesis. First, the LSP surface is
\emph{read-only}: it may emit diagnostics, hovers, code-action intents, virtual
documents, and protocol traces, but ``it must not directly mutate files''; any
future mutation must route through an explicit chain ending in a receipt
\citep{lspmax}. This realises the read-only posture that
\cref{sec:fw-lawstate} requires of a trustworthy observer. Second, the actuation
grammar at Layer~1 is a noun/verb CLI built on \code{clap-noun-verb}, in which a
filename is a noun and an annotated function is a verb---a small but real
demonstration that even the command surface is treated as a language with a
grammar.

\section{The \texorpdfstring{\maxm{*}}{max/*} Protocol Surface}
\label{sec:art-maxm}

Atop the standard LSP methods, \lspmax\ declares a custom \maxm{*} method family
(\code{lsp-max-protocol/src/custom\_methods.rs}) that exposes the law-state
runtime over JSON-RPC. \Cref{tab:maxm} groups the most load-bearing of these. The
surface is the concrete form of \cref{def:lawstate}(iii): the conformance state is
not an internal detail but a queryable protocol.

\begin{table}[t]
\centering
\caption[Representative \texttt{max/*} methods]{Representative methods of the
\maxm{*} surface \citep{lspmax}. The family exposes snapshots, three-valued
admission, receipts, repair plans, and release actuation over JSON-RPC.}
\label{tab:maxm}
\small
\begin{tabularx}{\textwidth}{@{}l Y@{}}
\toprule
\textbf{Method} & \textbf{Role} \\
\midrule
\maxm{snapshot}, \maxm{manifoldSnapshot} & capture server / full law-state snapshot \\
\maxm{conformanceVector} & query the three-valued conformance state at a snapshot \\
\maxm{admission} & admissibility gate over a law axis (\admitted/\refused/\unknownstat) \\
\maxm{refusal} & explicit refusal with rationale and receipt \\
\maxm{lawfulTransition} & assert that a phase transition is lawful \\
\maxm{explainDiagnostic}, \maxm{repairPlan} & explain a diagnostic; emit read-only repair steps \\
\maxm{receipt}, \maxm{replay} & retrieve a receipt; replay the event log against a snapshot \\
\maxm{releaseActuation} & actuate a release \emph{iff} the \code{ConformanceVector} admits it \\
\maxm{workspaceConformance}, \maxm{lsif} & aggregate workspace conformance; stream an LSIF graph \\
\bottomrule
\end{tabularx}
\end{table}

Note in particular \maxm{releaseActuation}: a release is gated on the conformance
vector admitting it, and \maxm{refusal} carries a rationale \emph{and} a receipt.
Admission and refusal are first-class, symmetric, and evidence-bound.

\section{The \texttt{ConformanceVector} and Three-Valued Law Logic}
\label{sec:art-cv}

The artifact's core type is the realisation of \cref{prin:three-valued}. It keeps
three disjoint sets of law axes and never coerces one into another
(\code{lsp-max-protocol/src/conformance.rs}).

\begin{lstlisting}[language=Rust,caption={The three-valued conformance state (abridged from \texttt{conformance.rs}).},label={lst:cv}]
pub struct ConformanceVector {
    pub admitted: Vec<LawAxis>,  // evidence present and valid
    pub refused:  Vec<LawAxis>,  // evidence present, violation confirmed
    pub unknown:  Vec<LawAxis>,  // admissibility NOT determinable
    pub score: Option<f64>,      // 100 * admitted / (admitted+refused+unknown)
    pub strict_mode: bool,       // do unknown axes block release?
    pub process_quality: Option<ConformanceResult>, // POWL conformance
    // bitmasks (not serialized) enforce mutual disjointness
}

impl ConformanceVector {
    pub fn all_admitted(&self) -> bool {
        self.refused.is_empty() && self.unknown.is_empty()
    }
    pub fn admits_release(&self) -> bool {
        self.refused.is_empty() && (!self.strict_mode || self.unknown.is_empty())
    }
}
\end{lstlisting}

Three properties deserve comment. First, \unknownstat\ is a peer of \admitted\ and
\refused, not the absence of a verdict; the three backing bitmasks are mutually
disjoint and an invariant check asserts it. Second, \code{admits\_release}
encodes a genuine policy choice: in \code{strict\_mode}, unknown axes block
release, so that ``not yet checked'' is treated as ``not yet permitted''. Third,
the law axes are an enumeration---\code{Protocol}, \code{Type}, \code{Fixture},
\code{Documentation}, \code{Release}, \code{Hook}, \code{Repair}, \code{Receipt},
\code{Security}, \code{Autopoiesis}, \code{Domain}, and an extensible
\code{Custom(String)}---so the same machinery scales from protocol conformance to
arbitrary domain law, which is exactly the extensibility \cref{ch:all-language}
will exploit. The \code{process\_quality} field, carrying a conformance result
from the process-mining engine, is the seam through which van der Aalst's
machinery enters the type.

\section{Receipts, Snapshots, and the Typestate Kernel}
\label{sec:art-receipts}

A claim of admission is worthless without something that cannot be cheaply
forged. \lspmax\ binds admission to a \emph{receipt} whose hash chains to its
predecessor (\code{lsp-max-protocol/src/core.rs}):

\begin{lstlisting}[language=Rust,caption={Content-addressed receipts form a hash chain (abridged from \texttt{core.rs}).},label={lst:receipt}]
pub struct Receipt {
    pub receipt_id: String,
    pub hash: String,                       // SHA-256 digest of the artifact
    pub prev_receipt_hash: Option<String>,  // None only for the genesis receipt
}
\end{lstlisting}

Each receipt's \code{prev\_receipt\_hash} points to the digest of the previous
receipt, forming a tamper-evident ledger whose genesis link is the only one
permitted to be \code{None}. The compositor additionally issues a
\code{CryptographicReceipt} chain (BLAKE3 hashing with Ed25519 signatures) so that
per-server contributions to a merged verdict remain attributable \citep{lspmax}.
A companion script validates the chain, requiring boundary markers
(\code{-----BEGIN RECEIPT-----}), a 64-hex digest, and a bounded status word, and
returning \refused\ when any are absent---so that the validator itself obeys the
three-valued discipline.

State evolution is governed by a typestate kernel
(\code{lsp-max-runtime/src/typestate/}) whose phases are monotonic---%
\code{Uninitialized} $\to$ \code{Initializing} $\to$ \code{Initialized} $\to$
\code{ShutDown} $\to$ \code{Exited}, with no back-edges---and whose trait exposes
a \code{validate} $\to$ \code{select} $\to$ \code{admit} $\to$ \code{receipt}
$\to$ \code{exit} $\to$ \code{replay} life cycle. The presence of \code{replay}
is significant: the runtime can reconstruct its state from the receipt ledger, so
the law-state is not merely asserted but \emph{re-derivable} from evidence---the
computational form of conformance checking's ``replay the log against the model''.

\section{The Anti-Cheat Canary and the Enforceable Constitution}
\label{sec:art-canary}

What makes \lspmax\ a \emph{law}-state runtime rather than a state machine with
aspirational documentation is that its constitution is enforced by a component of
the system itself: \code{anti-llm-cheat-lsp}, the ``diagnostic canary''. It runs
on \lspmax, exercises the LSP surface, and detects attempts to reintroduce the
prohibited base framework that \lspmax\ forks, to forge receipts or routes, or to
use the language of unbounded success \citep{lspmax}. The self-sealing property is
stated plainly in the constitution: \lspmax\ hosts the canary; the canary detects
regression; therefore \lspmax\ cannot silently regress \citep{lspmax}.

Detection is layered---raw-text scan, tree-sitter AST scan, manifest and
dependency-graph scan, claim scan over Markdown and agent reports, JSON-RPC
transcript scan, a receipt validator, a route-evidence checker, and a
claim-versus-proof checker---and every diagnostic must name the \emph{forbidden
implication} it prevents (for example, ``TestStdout $\Rightarrow$ Receipt'')
\citep{lspmax}. The diagnostic families span surface, authority, receipt, route,
mutation, test, version, and overclaim violations. Two enforcement mechanisms are
directly relevant to the thesis.

\paragraph{Bounded status and the refusal of victory language.} The constitution
forbids the words of triumph and enumerates the only permitted statuses---the
same vocabulary this thesis adopts in \cref{tab:status}. The canary's
overclaim family raises a diagnostic when forbidden terms appear, using an
Aho--Corasick automaton for linear-time classification \citep{lspmax}.

\paragraph{The ANDON gate.} A pre-tool-use hook runs \code{lsp-max-cli gate
check} before each shell, edit, or write action; exit~1 blocks the action while an
ANDON signal is set. The governing predicate is
\[
  \Lambda_{CD}(a) \;=\; \Lambda(a)\;\wedge\;\neg\,\exists\, d \in D_t :
  d.\mathit{law\_id} \in \mathcal{A}\ \wedge\ d.\mathit{severity}=\textsc{Error},
\]
where $\Lambda(a)$ is the base admissibility predicate, $D_t$ the live diagnostic
set, and $\mathcal{A}$ the governed law axes (\code{WASM4PM-*}, \code{ANTI-LLM-*},
\code{GGEN-*}) \citep{lspmax}. The gate state is published as a single byte in a
deterministically named file, so any process---including a subagent---can read it
in one syscall. The artifact is candid that subagent hooks do not cross session
boundaries, recording this as \openstat\ rather than \admitted: an honesty the
thesis takes as a model.

\section{OCEL Binding: Diagnostics as Object-Centric Events}
\label{sec:art-ocel}

The artifact's deepest point of contact with the paradigm of
\cref{sec:bg-pm}---and the strongest single piece of evidence for
\cref{hyp:substrate}---is that it does not merely \emph{borrow vocabulary} from
process mining; it \emph{runs} on it.

First, the constitution's laws are expressed as \emph{Declare} constraints---the
declarative, LTL-based process-modelling formalism from van der Aalst's
school---and mapped to diagnostics (\code{declare\_laws.rs}): the prohibition on
the base framework becomes \code{Absence}; ``a diagnostic must produce an OCEL
event'' becomes \code{Response}; ``a receipt must precede an admission'' becomes
\code{Precedence}; ``test output and a receipt cannot co-occur'' becomes
\code{NotCoexistence}; and ``an unknown axis cannot be succeeded by an admitted
one without a resolution event'' becomes \code{NotSuccession}. The artifact's
governance is, literally, a declarative process model checked for conformance.

Second, each diagnostic emission is recorded as an object-centric event
(\code{ocel.rs}). The object types are \code{CaseFile}, \code{DetectionCode},
\code{LawAxis}, \code{Receipt}, \code{Gate}, and \code{CodeSymbol}; a
\code{CheatDetected} event binds a file, a detection code, and a law axis, and a
\code{ScanComplete} event closes over the files scanned. The result is an OCEL
log of the agent's own conduct that any process-mining tool can read.

Third, conformance of that log is checked by the \code{wasm4pm} engine, which
returns a \code{GallVerdict}---\code{Fit\{fitness\}}, \code{Deviation\{fitness,
missing\}}, \code{Blocked\{reason\}}, or \code{Inconclusive\{reason\}}---whose four
cases map cleanly onto the bounded statuses and, notably, reserve
\code{Inconclusive} for the \unknownstat\ case. The engine's type-authority crate
further encodes OCEL axioms (disjoint event and object universes; mandatory
event-to-object cardinality; temporal continuity; type closure) and a family of
\emph{Oracle classes} (A8--A12) that recast adversarial behaviour---audit-log
tampering, temporal anomaly, causal violation, unknown-state collapse, cyclic
dependency---as \emph{violations of those axioms} \citep{lspmax}. Anti-cheat
detection is thereby lifted from pattern-matching to specification-derived
conformance checking.

\begin{table}[t]
\centering
\caption[Oracle classes as conformance violations]{The \code{wasm4pm} Oracle
classes recast cheating as object-centric axiom violations \citep{lspmax}. The
last row is the computational guardian of \cref{prin:three-valued}.}
\label{tab:oracle}
\small
\begin{tabularx}{\textwidth}{@{}l Y@{}}
\toprule
\textbf{Oracle} & \textbf{Formalised violation} \\
\midrule
A8 Audit-log tampering & an event with no causal predecessor \\
A9 Temporal anomaly & $e_1 \to e_2$ in causality but $t(e_1) \ge t(e_2)$ \\
A10 Causal violation & a mutation with no preceding observation \\
A11 Unknown-state collapse & \unknownstat $\to$ \{\admitted, \refused\} without a resolution event \\
A12 Cyclic dependency & a cycle in the causality graph \\
\bottomrule
\end{tabularx}
\end{table}

The artifact therefore exhibits, in miniature, the entire claim of this thesis:
a non-code ``language''---the agent's own stream of edits, claims, and
commands---is given an event-log abstraction (\cref{def:eventlog}), checked
against a declarative model on a conformance surface
(\cref{def:conformance-surface}), and reported in three-valued, receipt-bound,
read-only terms. What \cref{ch:all-language} does is widen the aperture from the
agent's conduct to natural language, law, mathematics, biology, music, and
business process. The mechanism is already present; only the domain changes.
