\chapter{The Fusion Architecture: LSP \texorpdfstring{$\times$}{x} MCP \texorpdfstring{$\times$}{x} A2A}
\label{ch:fusion}

\epigraph{A single Language Server can be re-used in multiple development tools,
which in turn can support multiple languages.}{\textit{Microsoft}, on LSP
\citep{lsp-official}}

\Cref{ch:background} established that LSP, MCP, and A2A are the same decoupling at
three layers; \cref{ch:framework} hypothesised that they are therefore strata of
one substrate (\cref{hyp:substrate}); \cref{ch:artifact} showed an artifact that
already runs the conformance stratum. This chapter composes the strata explicitly.
It assigns each protocol its role (\cref{sec:fus-lsp,sec:fus-mcp,sec:fus-a2a}),
shows how \lspmax\ occupies more than one stratum at once (\cref{sec:fus-stack}),
and traces a single conformance verdict through the whole stack
(\cref{sec:fus-conformance}).

\section{The Convergence Thesis, Concretely}
\label{sec:fus-convergence}

The fusion claim is not that the three protocols will be merged into one
specification. It is that they \emph{compose}: because all three are JSON-RPC
contracts with explicit capability negotiation
\citep{lsp-spec-318,mcp-spec-2025,a2a-spec}, a single agent can speak all three
without impedance, and the output of one becomes the input of another. An agent
discovered over A2A consults tools over MCP and reasons about language over a
generalised LSP; nothing in the three specifications obstructs this layering, and
their shared lineage and governance actively invite it
\citep{mcp-spec-2025,a2a-mcp,lf-aaif-press}. \Cref{fig:stack} renders the
resulting architecture.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  layer/.style={draw,rounded corners,minimum width=10.5cm,minimum height=10mm,align=center},
  >=Latex]
  \node[layer,fill=black!4] (agents) {\textbf{Agents} \;(human and autonomous)};
  \node[layer,fill=lawblue!10,below=4mm of agents] (a2a)
    {\textbf{A2A} --- coordination stratum \;\;(agent\,$\leftrightarrow$\,agent: Agent Cards, Tasks)};
  \node[layer,fill=lawblue!10,below=4mm of a2a] (mcp)
    {\textbf{MCP} --- context stratum \;\;(agent\,$\leftrightarrow$\,tools/data: resources, tools, prompts)};
  \node[layer,fill=lawblue!10,below=4mm of mcp] (lsp)
    {\textbf{LSP, generalised} --- semantic stratum \;\;(utterance\,$\leftrightarrow$\,meaning)};
  \node[layer,fill=black!4,below=4mm of lsp] (langs)
    {\textbf{Languages}\; code $\cdot$ natural $\cdot$ legal $\cdot$ math $\cdot$ bio $\cdot$ music $\cdot$ process};
  \draw[->] (agents) -- (a2a);
  \draw[->] (a2a) -- (mcp);
  \draw[->] (mcp) -- (lsp);
  \draw[->] (lsp) -- (langs);
  % cross-cutting conformance stratum
  \node[draw,rounded corners,fill=admitted!12,rotate=90,minimum width=4.9cm,
        minimum height=9mm,align=center,right=5mm of mcp.east,anchor=center,
        xshift=5mm] (conf)
    {\textbf{Conformance} (\lspmax)\\ three-valued $\cdot$ receipts $\cdot$ OCEL};
  \draw[decorate,decoration={brace,amplitude=4pt,mirror}]
    ($(a2a.north east)+(0.2,0)$) -- ($(lsp.south east)+(0.2,0)$);
\end{tikzpicture}
\caption[The fused substrate]{The fused substrate. A2A, MCP, and a generalised
LSP form three horizontal strata between agents and languages; \lspmax\ supplies a
cross-cutting conformance stratum (\cref{hyp:substrate}) that observes every layer
as an object-centric event log and reports three-valued, receipt-bound verdicts.}
\label{fig:stack}
\end{figure}

\section{LSP as the Semantic Substrate}
\label{sec:fus-lsp}

The lowest protocol stratum answers ``what does this utterance mean?''. Its
contribution to the fusion is the very property that made it successful for code:
it isolates language understanding behind a uniform contract so that the layers
above need not know how meaning is computed. Generalised per \cref{def:mml}, an
LSP server need not serve a programming language at all---LTeX already serves
natural language, and the YAML server serves a data format
\citep{ltex,yamlls}---and \lspmax\ extends the surface with the \maxm{*} family so
that ``meaning'' includes ``law-state'' (\cref{sec:art-maxm}). In the fused
architecture, every language domain of \cref{ch:all-language} appears as an
LSP-style server, integrated once and reachable by every agent above.

\section{MCP as the Context Substrate}
\label{sec:fus-mcp}

The middle stratum answers ``what may I consult, and what may I do?''. MCP's three
server primitives---resources, prompts, and tools \citep{mcp-spec-2025}---are the
natural envelope in which a generalised LSP server is presented to a model: a
language server's analyses become MCP \emph{resources}, its repair plans become
\emph{prompts}, and its actuations become \emph{tools}. The composition is
direct and already idiomatic, since MCP was itself shaped by LSP
\citep{mcp-spec-2025}. Concretely, \lspmax's read-only analyses (\maxm{conformanceVector},
\maxm{explainDiagnostic}) map to MCP resources, while its guarded actuations
(\maxm{releaseActuation}) map to MCP tools whose invocation the law-state runtime
may refuse. MCP thus carries conformance \emph{outward} to any host application
without that host needing to understand the conformance machinery.

\section{A2A as the Coordination Substrate}
\label{sec:fus-a2a}

The upper stratum answers ``which agents exist, and how do we combine?''. A2A's
Agent Card advertises capabilities for discovery \citep{a2a-spec,a2a-discovery};
in the fused architecture, ``produces receipt-bound conformance verdicts'' is
itself an advertisable capability, so that a coordinating agent can locate a
\emph{conformance authority} the way it locates any other skill. A2A's Task
lifecycle---\code{submitted}, \code{working}, \code{input-required},
\code{completed}, \code{failed}---is, read through \cref{ch:framework}, already an
event log: each state transition is an event with a timestamp and an actor. The
coordination stratum thus not only \emph{carries} conformance verdicts between
agents; its own activity is a language with a conformance surface, exactly as
\cref{sec:art-ocel} showed for an agent's stream of edits.

\section{The Composed Stack: \texorpdfstring{\lspmax}{lsp-max} as Mesh}
\label{sec:fus-stack}

\lspmax\ is instructive because it does not sit in a single stratum. Its Layer-2
surface is an LSP server (semantics); its \maxm{*} methods and \code{ConformanceVector}
are the conformance stratum; and its Layer-5 \emph{compositor} is a multi-server
mesh that fans a client request out to $N$ child servers and merges their
diagnostics, with the rule that refusals survive the merge
(``\code{REFUSED\_BY\_LAW} codes survive always'') \citep{lspmax}. This last
property is the architectural expression of \cref{prin:three-valued} at the mesh
level: aggregating many servers must not let an admission elsewhere overwrite a
refusal here. The compositor emits its own merge receipts and can render the
fan-out-then-merge flush as an OCEL event, so that even the act of composition is
minable. A mesh of conformance authorities, coordinated over A2A and consulted
over MCP, each understanding some language over LSP, is the fused substrate made
operational.

\section{Conformance Across the Fusion}
\label{sec:fus-conformance}

The point of fusing the strata is that a single verdict can travel the whole
height of \cref{fig:stack}. Consider a coordinating agent that must release a
regulated artifact. It discovers, over A2A, a conformance authority whose Agent
Card advertises the capability. It invokes that authority's analysis as an MCP
tool. The authority is, internally, an \lspmax\ mesh: it treats the artifact as
an object-centric event log (\cref{def:eventlog}), runs the conformance surface
over the relevant law axes (\cref{def:conformance-surface}), and returns a
\code{ConformanceVector}. If any axis is \refused, \maxm{releaseActuation} refuses
and emits a refusal receipt; if axes are \unknownstat\ and the policy is strict,
release is withheld rather than granted; only if the vector admits release does
the coordinating agent proceed---carrying the receipt as evidence that other
agents, over A2A, can in turn verify.

Two features of this flow are worth isolating as the fusion's distinctive
contributions. First, \emph{accountability composes with reach}: the same
capability-negotiated, JSON-RPC machinery that lets the agent reach further also
lets a verdict and its receipt travel back, so expansion of reach need not
outrun expansion of accountability---the deficiency P2 of \cref{sec:problem}.
Second, \emph{the verdict is portable}: because the receipt is content-addressed
and chained (\cref{lst:receipt}), an admission produced deep in an \lspmax\ mesh
remains checkable after it has crossed an MCP tool boundary and an A2A task
boundary. The fusion does not merely connect systems; it lets a single,
three-valued, evidence-bound judgement remain meaningful across every connection
it crosses. Whether this portability holds under adversarial conditions at scale
is, properly, \candidate\ rather than \admitted, and \cref{ch:discussion} returns
to it.
