\chapter{Conformance Checking for All Language}
\label{ch:all-language}

\epigraph{In the case of LTeX, the ``language'' is not a single natural language,
but all natural languages.}{\textit{LTeX documentation} \citep{ltex}}

This chapter discharges contributions~C2 and~C5. It states the criterion that
makes a domain a servable language (\cref{sec:al-criterion}), then works six
domains through the pipeline of \cref{fig:pipeline}---natural and legal language
(\cref{sec:al-nl,sec:al-legal}), mathematics (\cref{sec:al-math}), biological
sequence (\cref{sec:al-bio}), music (\cref{sec:al-music}), and business process
(\cref{sec:al-bpm})---adds the reflexive case of an agent's own conduct
(\cref{sec:al-reflexive}), and closes with the object-centric bridge and the single
conformance functor that unify them (\cref{sec:al-bridge,sec:synthesis-functor}).
Per \cref{sec:m-eval}, these are illustrative scenarios:
they demonstrate feasibility by construction and expose limits; they are graded
\candidate, not \admitted, and field validation is \openstat.

\section{The Criterion: When Is a Domain a Language?}
\label{sec:al-criterion}

\Cref{def:eventlog} gives the test. A domain is a servable language when its use
admits an object-centric event-log abstraction: its activities can be recorded as
timed events referring to typed objects. Given that, \cref{def:conformance-surface}
supplies a conformance surface, and the artifact of \cref{ch:artifact} supplies
the machinery---three-valued verdicts, receipts, OCEL events---without
modification, because the law axes are extensible (the \code{Custom} variant of
\cref{lst:cv}). The criterion is deliberately restrictive: a domain whose use
cannot be observed as events (a purely private, untraced mental act) falls
outside the claim, and saying so is what keeps ``all language'' from meaning
``everything''. \Cref{tab:domains} previews the six demonstrations.

\begin{table}[t]
\centering
\caption[Six languages under one conformance surface]{Six domains as servable
languages. Each admits an event-log abstraction (activity over typed objects) and
therefore a conformance surface; existing tooling shows the semantic stratum is
already partly built.}
\label{tab:domains}
\scriptsize
\begin{tabularx}{\textwidth}{@{}l Y Y Y l@{}}
\toprule
\textbf{Domain} & \textbf{Event (activity)} & \textbf{Typed objects} & \textbf{Example law axis} & \textbf{Semantic-stratum tooling} \\
\midrule
Code & edit, parse, build & symbol, module, file & type safety, API stability & LSP servers \citep{lsp-official} \\
Natural language & draft, revise, translate & sentence, term, claim & grammar, style, consistency & LTeX, textLSP \citep{ltex,textlsp} \\
Legal / regulatory & draft clause, amend, cite & clause, obligation, party & obligation consistency, citation validity & emerging \\
Mathematics & state, step, justify & proposition, lemma, term & logical validity, acyclic dependency & proof assistants \\
Biological seq. & transcribe, edit, annotate & base, gene, variant & reading-frame integrity, reference conformance & bioinformatics validators \\
Music & state, transpose, vary & note, motif, voice & voice-leading, key/metre conformance & notation formats \\
Business process & execute, hand off & case objects, resource & control-flow conformance & process mining \citep{vdaalst-book-2016} \\
\bottomrule
\end{tabularx}
\end{table}

\section{Natural Language}
\label{sec:al-nl}

Natural language is the domain where the semantic stratum is most mature outside
code. LTeX serves grammar and style over LSP for ``all natural languages
supported by LanguageTool'' \citep{ltex}; textLSP and similar servers add prose
checks \citep{textlsp}. The process view adds what these lack: a memory of
behaviour. The \emph{events} are drafting and revision acts; the \emph{objects}
are sentences, terms, and claims; a model $M_L$ encodes constraints such as
``a defined term is used consistently'' or ``a claim asserted is not later
contradicted''. Conformance checking then locates contradiction or terminological
drift as a \refused\ axis, with the offending events as the diagnostic's witness,
and marks as \unknownstat\ those claims whose truth it cannot assess---refusing,
per \cref{prin:three-valued}, to label the unchecked as correct. This is the
direct generalisation of the artifact's own claim-versus-proof checker
(\cref{sec:art-canary}) from agent reports to arbitrary prose.

\section{Legal and Regulatory Text}
\label{sec:al-legal}

Law is language whose entire purpose is conformance, which makes it the domain
where the thesis bites hardest---and the one the artifact's ``affidavit'' lineage
(\cref{ch:artifact}) already gestures at, through receipts that function as
cryptographic attestations. The \emph{events} are the drafting, amendment, and
citation of provisions; the \emph{objects} are clauses, obligations, parties, and
referenced statutes, with exactly the object-to-object relationships OCEL~2.0 was
extended to express (an obligation \emph{borne by} a party, a clause
\emph{amending} another) \citep{ocel2-2024}. A model $M_L$ encodes obligation
consistency (no clause both requires and forbids the same act), citation validity
(every reference resolves to an in-force provision), and temporal coherence
(amendments respect effective dates---precisely the A9 ``temporal anomaly'' axiom
of \cref{tab:oracle}). A conformance verdict over a contract or regulation is
then a receipt-bound, three-valued document: \admitted\ obligations, \refused\
contradictions, and \unknownstat\ provisions whose compliance depends on facts
not in evidence. The receipt is the natural digital analogue of a sworn
attestation.

\section{Mathematics and Formal Proof}
\label{sec:al-math}

Mathematical proof is language with the strictest conformance model of all:
logical validity. Proof assistants already expose their language over LSP, so the
semantic stratum exists. The \emph{events} are the statement of propositions and
the application of inference steps; the \emph{objects} are propositions, lemmas,
definitions, and terms, related by \emph{depends-on}. The model $M_L$ is the
proof calculus itself, and conformance checking is proof checking: a step that
does not follow is \refused; a lemma invoked before it is established is an A10
``causal violation'' (a mutation---use---preceding its observation---proof); a
circular dependency among lemmas is the A12 ``cyclic dependency'' of
\cref{tab:oracle}. That the artifact's Oracle classes, designed to catch agent
cheating, map without strain onto proof pathologies is evidence for the unity the
thesis claims.

\section{Biological Sequence}
\label{sec:al-bio}

A genome is a language over a four-letter alphabet with rich grammar. The
\emph{events} are transcription, editing, and annotation operations; the
\emph{objects} are bases, codons, genes, variants, and regions; the model $M_L$
encodes reading-frame integrity, conformance to a reference sequence, and
constraints on permissible edits. Conformance checking flags a frame-shifting
edit as \refused, a variant of unknown significance as \unknownstat---a domain
that has independently institutionalised \cref{prin:three-valued} under the very
name ``variant of uncertain significance''---and a reference-matching region as
\admitted. The semantic-stratum tooling here is bioinformatics validation rather
than an LSP server today; the contribution of the thesis is to observe that the
\emph{same} conformance surface serves it, and that an LSP-style server is a
natural packaging.

\section{Music and Multimodal Notation}
\label{sec:al-music}

A score is notation with grammar (key, metre, voice-leading). The \emph{events}
are the statement, transposition, and variation of material; the \emph{objects}
are notes, motifs, voices, and bars; the model $M_L$ encodes voice-leading rules,
key conformance, and metric consistency. Conformance checking locates parallel
fifths or a metric overflow as \refused, and leaves stylistically open choices
\unknownstat. Music is included not for its economic weight but because it
stresses the criterion of \cref{sec:al-criterion}: it shows the event-log
abstraction applies to a non-textual, time-structured, aesthetic language, and
that ``law'' need not mean ``regulation'' to be conformance-checkable.

\section{Business Process: The Return Home}
\label{sec:al-bpm}

The sixth domain is the one process mining began with, and including it closes the
loop. Here the \emph{events} are activity executions and hand-offs; the
\emph{objects} are the multiple case objects whose entanglement motivated OCEL in
the first place (\cref{sec:bg-ocel}); the model $M_L$ is a Petri net or a POWL
model; and conformance checking is conformance checking in van der Aalst's
original sense, computed by the \code{wasm4pm} engine and returned as a
\code{GallVerdict} (\cref{sec:art-ocel}). The thesis adds nothing to this domain
that the field did not already have---which is the point. Business process is the
\emph{calibration} case: because the artifact's machinery reproduces standard
object-centric conformance here, its application to the other five domains is a
generalisation of a working method, not an analogy to a foreign one.

\section{The Reflexive Case: An Agent's Own Conduct}
\label{sec:al-reflexive}

The seventh language is the one the artifact already serves: an agent's own stream
of edits, claims, and commands (\cref{sec:art-ocel}). The \emph{events} are tool
calls, edits, and assertions; the \emph{objects} are the artifact's OCEL types
(\code{CaseFile}, \code{DetectionCode}, \code{LawAxis}, \code{Receipt}, \code{Gate},
\code{CodeSymbol}); the model $M$ is the Declare-encoded constitution; and
conformance is the canary's verdict, emitted as \code{CheatDetected} events and
checked by the engine's \code{GallVerdict} (\cref{sec:art-ocel}). Two features
distinguish this case. First, where business process (\cref{sec:al-bpm}) calibrates
against van der Aalst's original setting, the reflexive case calibrates against an
actually-running implementation, so its status is \admitted-by-dogfood rather than
\candidate. Second, it is self-applying: the conformance surface is policed by the
same machinery it governs, the self-sealing property of \cref{sec:art-canary}. The
reflexive case is therefore both the seventh demonstration and the existence proof
that the other six are not merely conceivable---the functor of
\cref{thm:conf-functor} is inhabited on at least one object, witnessed in code.

\section{The Object-Centric Bridge}
\label{sec:al-bridge}

What lets one surface serve six domains is the object-centric event log. A flat,
single-case log would force each domain into a single ``case'' notion---one
sentence, one proof, one gene---and suffer the convergence and divergence
pathologies of \cref{sec:bg-ocel}: a legal obligation borne by several parties, a
lemma used in several proofs, a motif recurring in several voices would each be
mismodelled. OCEL~2.0's object-to-object and qualified relationships are exactly
what these domains require \citep{ocel2-2024}. This is the precise sense in which
van der Aalst's ``no AI without PI'' \citep{noai-nopi-2025} becomes, for this
thesis, \emph{no conformance for all language without an object-centric
abstraction}: the bridge that carries conformance from business process to all
language is the object-centric event log, and it is load-bearing. Formally, this
unity is \cref{cor:one-functor}: the seven domains are objects of one category and
conformance is one functor (\cref{thm:conf-functor}). The unity is summarised by a
single observation: across all seven domains, the artifact's machinery changed not
at all; only the law axes and the model $M_L$ changed. That
invariance, demonstrated by construction, is the evidence offered for
\cref{hyp:substrate}'s semantic-plus-conformance claim---offered, per
\cref{sec:m-eval}, as \candidate.
