\chapter{Conclusion and Future Work}
\label{ch:conclusion}

\epigraph{Optimise for admitted work.}{\textit{\lspmax}, \texttt{AGENTS.md}}

\section{Summary of Contributions}
\label{sec:conc-summary}

This thesis argued that the convergence of three protocols---the Language Server
Protocol generalised beyond code, the Model Context Protocol, and the
Agent-to-Agent protocol---under a conformance stratum constitutes a phase
transition in machine-mediated language, and that the discipline founded by Wil van
der Aalst supplies the missing surface that makes the expansion trustworthy. The
six contributions stand, with bounded status, as follows.

\begin{itemize}[leftmargin=2.2em]
  \item[\textbf{C1}] The unifying account of LSP, MCP, and A2A as one decoupling
  performed at three layers (\cref{ch:background}), formalised by the Decoupling
  Theorem (\cref{thm:decoupling}). \admitted\ analytically.
  \item[\textbf{C2}] The event-log criterion for when a domain is a servable
  language (\cref{def:eventlog}). \admitted\ as a criterion.
  \item[\textbf{C3}] The phase-transition model of convergence, with the
  latent-heat-of-standardisation formalism and a from-first-principles derivation of
  Clausius--Clapeyron and a Landau jump (\cref{ch:math,ch:framework}). \admitted\ as
  a model; its 2030 instantiation is a \forecast.
  \item[\textbf{C4}] The artifact, \lspmax, analysed as a law-state runtime whose
  \code{ConformanceVector} keeps \admitted/\refused/\unknownstat\ strictly distinct
  and whose diagnostics are object-centric events (\cref{ch:artifact}). \admitted\
  for the four defining properties, with \openstat\ sub-items.
  \item[\textbf{C5}] The demonstration of conformance-for-all-language across six
  domains, unified by the object-centric bridge (\cref{ch:all-language}).
  \candidate.
  \item[\textbf{C6}] The 2030 roadmap and the governance-and-risk analysis, with
  forecasts flagged (\cref{ch:vision}). \candidate, with \openstat\ later stages.
\end{itemize}

The through-line is a single inversion. Van der Aalst's dictum holds that there is
``no AI without PI''---no trustworthy artificial intelligence without the process
intelligence that object-centric mining provides \citep{noai-nopi-2025}. This
thesis advanced its contrapositive across language: \emph{there is no conformance
for all language without an object-centric event abstraction}. The protocols
provide reach; process mining provides the account of whether that reach was
lawfully used; and the object-centric event log is the bridge that carries
conformance from business process---where van der Aalst built it---to natural
language, law, mathematics, biological sequence, music, and back.

\section{Future Work}
\label{sec:conc-future}

Four directions follow directly. \emph{Empirical instantiation}: build and exercise
generalised LSP-plus-conformance servers for two or three of the six domains, emit
their behaviour as OCEL~2.0, and measure conformance against domain models---moving
C5 from \candidate\ toward \admitted. \emph{Cross-domain law axes}: investigate
whether a shared vocabulary of law axes (\cref{lst:cv}'s extensible enumeration)
can span domains without collapsing their differences, the prerequisite for
roadmap stage~S4. \emph{Receipt portability under adversarial composition}: test
whether a verdict's receipt survives crossing MCP and A2A boundaries when
participants are adversarial, the open question left by \cref{sec:fus-conformance},
in dialogue with the cryptographic-binding and verifiable-delegation literature
\citep{zhou26gov,prakash26aip}. \emph{The guardian's own guarantees}: formalise the
trust one may place in a conformance stratum given finite oversight capacity
\citep{turan26} and the limits of proof-of-guardrail \citep{jin26guardrail}---the
question of whether the guardian guards itself.

\section{Closing}
\label{sec:conc-closing}

A phase transition is not louder than what precedes it; it is discontinuous with
it. Water at the boiling point gives no warning in its temperature that, one unit
of latent heat later, it will occupy a thousand times the space
\citep{nationalboard-steam,latentheat-openstax}. The standardisation of how
machines mean, consult, and coordinate has been, for a decade, such a plateau:
much effort, little visible change. This thesis has argued---bounding the claim, and
declining the language of triumph---that the plateau is the latent heat of a
transition, not its absence; that the transition's signature will be conformance,
not merely connection; and that the discipline prepared to read the new volume of
language is the one that already learned to read behaviour as an event log. Whether
the threshold is crossed by 2030 is a \forecast. That the surface for reading it is
buildable is \admitted: it has been built, in part, in the artifact this thesis
studied. The remainder is \openstat---which is the only honest place for a thesis
about conformance to end.
